Bayesian inference of prehistoric demography from multiple proxies: a case study from the Swiss plateau

### 2022-05-19 manuscript/paper.Rmd
Edits and suggestions to introduction

## Prehistoric demography as object of research

The study of population size and population density as important demographic factors has a long history in archaeology. Although questions of population extent and distribution have been dominant fields since the beginning of archaeological research, the lack of access to this information, data on it, or methodological tools for the use of proxies has long made the statements made on this topic for prehistoric times vague and superficial [@hassan_demographic_1981], and they were often based on uncritical transfer of ethnographic parallels.

Gordon Childe [-@childe_man_1936] was one of the first to stress the importance of estimating the density and characteristics of prehistoric populations and sought to examine the role of human populations in the course of cultural evolution [@hassan_demographic_1981].

Attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946]. Cook and Heizer (1965, 1968) and Naroll (1962) provided rules derived from ethnographic contexts for estimating the size of prehistoric populations. Cook and Heizer (1966) made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981].

At the end of the 1960s, however, with the emergence of the processual archaeology, such investigations became increasingly popular. It was the increased focus on human-environment relations paradigm that increased the emphasis on population studies, especially those dealing with population ecology [@hassan_demographic_1981].

After research interest turned increasingly to other aspects with the emergence of post-process trends, a strong boom in the study of population sizes can be observed especially since the 2010s [@riede_climate_2009 and others in that special issue of Human Biology, or even earlier, @shennan_population_2000]. This is probably not at least due to the new shift towards the study of human-environment relationships, for which an assessment of the size of human populations is indispensable. This is true especially for an evaluation of the human impact on the natural environment. Another reason is certainly the emergence of the 'dates as data' approach [@rick_dates_1987], which only really took off with the publications around the London UCL group [@shennan_regional_2013], but also others []. Moreover, Kintigh et al. [-@kintigh_grand_2014] listed human influence, dominance and population size and growth as one of the key elements where methodological improvements are necessary in the article on the Grand Challenges for archaeology, in the area of 'human-environment interactions'. Methodology continues to evolve, but we believe that essential problems underlying summation calibration cannot be solved by methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].

## Issues associated with existing approaches

>>>>

## Prehistoric demography

Prehistorians have long recognised demography as a fundamental force in human cultural evolution [@childe_man_1936]. However, despite decades of interest in the demographic dynamics of prehistoric societies, concrete estimates of population size and population density before written records remain elusive. Though the archaeological record provides multiple possible demographic proxies [@hassan_demographic_1981], lack of access to this data, and a lack of methodological tools for turning it into quantitative estimates, has often left the conclusions drawn from it vague and superficial [@hassan_demographic_1981]. As a result, 'expert estimates' transferred from ethnographic parallels have often taken the place of direct inference from archaeological evidence.<!--REF? Turchin? Morris?-->

Early attempts to estimate the size and density of prehistoric populations from archaeological remains were made by Hack [-@hack_changing_1942], Colton [-@colton_prehistoric_1949], Frankfort [-@frankfort_town_1950] and Cook [-@cook_reconsideration_1946]. Cook and Heizer (1965, 1968) and Naroll (1962)<!--Are these in the bibliography?--> provided rules derived from ethnographic contexts for estimating the size of prehistoric populations. Cook and Heizer (1966) also made the first attempt to estimate the population growth rate during the Neolithic [@hassan_demographic_1981]. These approaches became increasingly common after the 1960s, with the emergence of processual archaeology and its focus on human–environment relations and population ecology [@hassan_demographic_1981].

Following a spell out of fashion with the post-processual turn, prehistoric demography has experienced a revival in recent years as part of the cultural evolution paradigm [@shennan_population_2000; @riede_climate_2009 and others in same issue]. Part of this resurgence can be explained by a renewed interest in human–environment relations and human impact on the natural environment, which necessarily requires an assessment of population size. Kintigh et al. [-@kintigh_grand_2014], for example, list human influence, dominance, population size, and population growth amongst their "grand challenges" for archaeology in the 21st century.
But methodological advances have certainly also played a role. In the last decade, the 'dates as data' technique [@rick_dates_1987], which uses the frequency of radiocarbon dates as a proxy for population dynamics, was taken up and significantly developed in a series of publications from the UCL group [e.g. @shennan_regional_2013], and has since been widely applied to archaeological contexts around the world [@crema_review_2022].

The 'dates as data' or summed radiocarbon approach has contributed greatly to our understanding of prehistoric demography, but it is not without its critics [@attenbrow_hiscock_2015, @carleton_groucutt_2020]. While the methodology continues to evolve and to an extent accommodate these critiques [@crema_review_2022], there are certain essential problems that cannot be solved with methodological refinements in this area alone [@french_manifesto_2021, @schmidt_approaching_2021].<!-- [JR] What problems? -->

## Evaluation of existing approaches
